Development of restored dike grasslands to reference needs more than four years and can hardly modified by local site conditions

### 2023-02-02 manuscript.Rmd
Update manuscript.Rmd

The seed mixture-substrate combination were tested with an experiment on a dike of the River Danube in SE Germany, which was established in March 2018 (Figure \@ref(fig:figure-map)); 314 m above sea level; WGS84 (lat/lon), 48.83895, 12.88412). The climate is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [@hari2020; @cdc2022temp; @cdc2022prec], and three minor floods which did not reach the plots [Appendix Table @ref(tab:flood-weather), @lfu2021waterlevel; @lfu2021discharge]. Regional sand (0--4 mm) was used to lean the substrate and the soil was taken from removed agricultural soil of a dike construction site near Steinkirchen. The substrates were mixed with a big roller and an excavator prepared the plots. The target vegetation were lowland hay meadows [EUNIS code: R22, @Chytry.2020; Arrhenatherion elatioris, EuroVegChecklist: CM01A, @mucina2016] and semi-dry calcareous grassland (R1A; Cirsio-Brachypodion pinnati, DA01B). The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds. From the species pools, 20 species were randomly selected for each plot. The seed mixtures always contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The community-weighted means (CWM) of functional traits differed between the seed mixture types (Table \@ref(tab:traits)) . The management started with a cut without hay removal and a mowing height of 20 cm in August 2018, followed by 'normal' deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was seeded as a nursery grass to provide safe sites under drought conditions.

We used 288 plots of the size 2.0 m × 2.0 m which were distributed over the north and south exposition, halfway up the slopes of the dike and arranged in six blocks (= replicates). The experiment used a split plot design combined with randomized complete block design (Figure \@ref(fig:figure-map)). The split plot was caused by testing all 24 treatments on both sides of the dike. We tested three sand admixtures (0%, 25% and 50%) and tow soil depths (15 vs. 30 cm) of practical interest [@KleberLerchbaumer.2017] and two . Below the substrate, a drainage layer of 5 cm consisting of gravel (0--16 mm) was installed. We tested soil samples of the three substrates on both expositions by mixing several subsamples. The sand admixture changed the soil texture, increased the C/N ratio, but reduced the ratio of Calciumcarbonat, and hardly changed the pH (Table @ref(tab:substrate)). The two seed mixture types were tested with the normal seeding density and the increased density recommended for slopes [4 vs. 8 g m^-2^ @Kiehl.2010; @KleberLerchbaumer.2017]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. Establishment rates of species (Appendix Table \@ref(tab:establishment)).

@@

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists of each plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used 4 chains, a thinning rate of 2, 5,000 iterations for warm-up and 10,000 in total. For evaluating the computation, MCMC chain convergence was checked with trace plots and by the R-hat values and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive check was done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019].

@@

Data, code and the whole model specification and evaluation is stored on Zenodo and in an easily accessible a scroll-down document on GitHub [cite Zenoto]<!--# Cite Zenodo--> There, the sections are referenced to the Bayesian analysis reporting guidelines [BARG, @kruschke2021]. All analyses were performed in R [Version 4.2.2, @base] with the functions 'brm' from the package 'brms' [@brms] for model calculation, several functions from 'brms' and 'bayesplot' [@bayesplot] for model evaluation and 'metaMDS' from 'vegan' for the ordination [@vegan].

@@

We could identify a statistically clear positive effect of sand addition on the persistence of sown species and on the recovery rate (Figure \@ref(fig:figure-effects)). Furthermore, sown hay meadows were clearly closer to their reference plots than sown dry grasslands were to their references. The posterior distributions which are shown in the interactions plots which separates exposition and survey year indicate for all three responses a strong positive development after one year with slow increase in following years (Figures \@ref(fig:figure-persistence); \@ref(fig:figure-fcs); \@ref(fig:figure-recovery)) . Both expositions have similar trends but for all responses the values are clearly lower on the south exposition. Persistence values are more than 50% higher on the north exposition<!--# insert further percentages--> (Figure \@ref(fig:figure-persistence)). The effect of restoration treatments is in the interaction plots not clear. Hay meadow seed mixture led to a closer development to hay meadow references and almost reached on the north exposition the area of the standard deviation of reference sites (Figure \@ref(fig:figure-recovery)). Though, the sand admixture had no clear effect on the development. Persistence of both seed mixture types seems to be slightly positive affected <!--# insert percentage-->by sand admixture on the north exposition (Figure \@ref(fig:figure-persistence)).

@@

MB is funded by a doctoral scholarship of the German Federal Environmental Foundation (DBU) (No. 20021/698). The establishment of the experiment and the vegetation surveys were financed by the WIGES GmbH in the years 2018--2020. <!--# funding number of WIGES -->

@@

```{r tab.id = "substrate", ft.align = "left",tab.cap = "Soil of the three substrates was analyzed for the fraction < 2mm fraction. Soil texture was classified according to the Bodenkundliche Kartieranleitung [@BodenkundlKartieranl.2005]. The pH was measured in CaCl2 solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in CaCl2 extract. For calculating CaCO3, a subsample was annealed at 550 °C and the measured C amount calculated with 8.33. To calculate total N and the C/N ratio, a subsample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand."}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(i = 1, j = 10, part = "header", value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))) %>%
  compose(i = 1, j = 11, part = "header", value = as_paragraph("K", as_sub("2"), "O")) %>%
  compose(i = 1, j = 12, part = "header", value = as_paragraph("Mg", as_sup("2+"))) %>%
  compose(i = 1, j = 14, part = "header", value = as_paragraph("CaCO", as_sub("3"))) %>%
  compose(i = 2, j = 10, part = "header", value = as_paragraph("mg 100g", as_sup("−1"))) %>%
  compose(i = 2, j = 11, part = "header", value = as_paragraph("mg 100g", as_sup("−1"))) %>%
  compose(i = 2, j = 12, part = "header", value = as_paragraph("mg 100g", as_sup("−1"))) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

<!--# Cite BayernAtlas-->

```{r figure-map, fig.asp = 1.2854, fig.width = 5, fig.cap = "The experiment was located on a dike at river Danube in SE Germany.  The 288 plots were allocated in six blocks (red squares on upper photo) and on the north and south exposition (central photo). Four treatments were conducted (Sand admixture, substrate depth, seeding density, seed mixture types H and D). The western half of a block was with a shallow substrate depth for the blocks and within this half the substrates. The photo on the bottom shows one block from the side after four years (Photo: Markus Bauer)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_1_map.png")
  )
```

@@

```{r fig.id = "figure-effects", fig.asp = 0.3030, fig.width = 6.49, fig.cap = "The posterior density distributions (grey) of the treatments calculated over all four surveyed years. Furthermore, the median and the 66% and 95% credible intervals are shown. We evaluated the persistence of sown species (A), the Favourable Conservation Status (FCS; B) and the successional distance to reference (recovery; C). Note that the zero lines indicate that both levels have equal values. This means that hay meadows are closer to their reference than dry grasslands  (C)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_2_800dpi_16.5x5cm.tiff")
  )
```

@@

```{r fig.id = "figure-fcs", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The Favourable Conservation Status (FCS) of a dike experiment. The substrates had different sand admixtures and different seed mixture types were sown. The zero line indicate that target and non-target species are balanced. Positive values indicate that there are more target species.  Shown are the posterior distributions, their medians and 95 % credible intervals derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_4_fcs_epred_800dpi_16.5x8cm.tiff")
  )
```

>>>>

The seed mixture-substrate combination were tested with an experiment on a dike of the River Danube in SE Germany, which was established in March 2018 (Figure \@ref(fig:figure-map)); 314 m above sea level; WGS84 (lat/lon), 48.83895, 12.88412). The climate is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [@hari2020; @cdc2022temp; @cdc2022prec], and three minor floods which did not reach the plots [Appendix Table @ref(tab:flood-weather), @lfu2021waterlevel; @lfu2021discharge]. Regional sand (0--4 mm) was used to lean the substrate and the soil was taken from removed agricultural soil of a dike construction site near Steinkirchen. The substrates were mixed with a big roller and an excavator prepared the plots. The target vegetation were lowland hay meadows [EUNIS code: R22, @Chytry.2020; Arrhenatherion elatioris, EuroVegChecklist: CM01A, @mucina2016] and semi-dry calcareous grassland (R1A; Cirsio-Brachypodion pinnati, DA01B). The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds. From the species pools, 20 species were randomly selected for each plot. The seed mixtures always contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The community-weighted means (CWM) of functional traits differed between the seed mixture types (Table \@ref(tab:traits))<!--# describe difference --> . The management started with a cut without hay removal and a mowing height of 20 cm in August 2018, followed by 'normal' deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was seeded as a nursery grass to provide safe sites under drought conditions.

We used 288 plots of the size 2.0 m × 2.0 m which were distributed over the north and south exposition, halfway up the slopes of the dike and arranged in six blocks (= replicates). The experiment used a split plot design combined with randomized complete block design (Figure \@ref(fig:figure-map)). The split plot was caused by testing all 24 treatments on both sides of the dike. We tested three sand admixtures (0%, 25% and 50%) and tow soil depths (15 vs. 30 cm) of practical interest [@KleberLerchbaumer.2017] and two . Below the substrate, a drainage layer of 5 cm consisting of gravel (0--16 mm) was installed. We tested soil samples of the three substrates on both expositions by mixing several sub-samples. The sand admixture changed the soil texture, increased the C/N ratio, but reduced the ratio of Calciumcarbonat, and hardly changed the pH (Table @ref(tab:substrate)). The two seed mixture types were tested with the normal seeding density and the increased density recommended for slopes [4 vs. 8 g m^-2^ @Kiehl.2010; @KleberLerchbaumer.2017]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. Establishment rates of species (Appendix Table \@ref(tab:establishment)).

@@

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists of each plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used 4 chains, a thinning rate of 2, 5000 iterations for warm-up and 10,000 in total. For evaluating the computation, MCMC chain convergence was checked with trace plots and by the R-hat values and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive check was done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019].

@@

Data, code and the whole model specification and evaluation is stored on Zenodo and in an easily accessible a scroll-down document on GitHub [!cite Zenodo!]<!--# Cite Zenodo--> There, the sections are referenced to the Bayesian analysis reporting guidelines [BARG, @kruschke2021]. All analyses were performed in R [Version 4.2.2, @base] with the functions 'brm' from the package 'brms' [@brms] for model calculation, several functions from 'brms' and 'bayesplot' [@bayesplot] for model evaluation and 'metaMDS' from 'vegan' for the ordination [@vegan].

@@

We could identify a statistically clear positive effect of sand addition on the persistence of sown species and on the recovery rate (Figure \@ref(fig:figure-effects)). Furthermore, sown hay meadows were clearly closer to their reference plots than sown dry grasslands were to their references. The posterior distributions which are shown in the interactions plots which separates exposition and survey year indicate for all three responses a strong positive development after one year with slow increase in following years (Figures \@ref(fig:figure-persistence); \@ref(fig:figure-fcs); \@ref(fig:figure-recovery)) . Both expositions have similar trends but for all responses the values are clearly lower on the south exposition, e.g., persistence values are on average more than 46% higher on the north exposition (Figure \@ref(fig:figure-persistence)). The effect of restoration treatments is in the interaction plots not clear. Hay meadow seed mixture led to a closer development to hay meadow references and almost reached on the north exposition the area of the standard deviation of reference sites (Figure \@ref(fig:figure-recovery)). Though, the sand admixture had no clear effect on the development. Persistence of both seed mixture types seems to be slightly positive affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:figure-persistence)).

@@

MB is funded by a doctoral scholarship of the German Federal Environmental Foundation (DBU) (No. 20021/698). The establishment of the experiment and the vegetation surveys were financed by the WIGES GmbH in the years 2018--2020 (No. 80 002 312).

@@

```{r tab.id = "substrate", ft.align = "left",tab.cap = "Soil of the three substrates was analyzed for the fraction < 2mm fraction. Soil texture was classified according to the 'Bodenkundliche Kartieranleitung' [@BodenkundlKartieranl.2005]. The pH was measured in CaCl2 solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in CaCl2 extract. For calculating CaCO3, a subsample was annealed at 550 °C and the measured C amount calculated with 8.33. To calculate total N and the C/N ratio, a subsample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand."}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(i = 1, j = 10, part = "header", value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))) %>%
  compose(i = 1, j = 11, part = "header", value = as_paragraph("K", as_sub("2"), "O")) %>%
  compose(i = 1, j = 12, part = "header", value = as_paragraph("Mg", as_sup("2+"))) %>%
  compose(i = 1, j = 14, part = "header", value = as_paragraph("CaCO", as_sub("3"))) %>%
  compose(i = 2, j = 10, part = "header", value = as_paragraph("mg 100g", as_sup("−1"))) %>%
  compose(i = 2, j = 11, part = "header", value = as_paragraph("mg 100g", as_sup("−1"))) %>%
  compose(i = 2, j = 12, part = "header", value = as_paragraph("mg 100g", as_sup("−1"))) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

```{r figure-map, fig.asp = 1.2854, fig.width = 5, fig.cap = "The experiment was located on a dike at river Danube in SE Germany.  The 288 plots were allocated in six blocks (red squares on upper photo) and on the north and south exposition [central photo; both aerial photos: @bayerischevermessungsverwaltung2023]. Four treatments were conducted (Sand admixture, substrate depth, seeding density, seed mixture types H and D). The western half of a block was with a shallow substrate depth for the blocks and within this half the substrates. The photo on the bottom shows one block from the side after four years (Photo: Markus Bauer)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_1_map.png")
  )
```

@@

```{r fig.id = "figure-effects", fig.asp = 0.3030, fig.width = 6.49, fig.cap = "The posterior density distributions (grey) of the treatments calculated over all four surveyed years. Furthermore, the median and the 66% and 95% credible intervals are shown. We evaluated the persistence of sown species (A), the Favorable Conservation Status (FCS; B) and the successional distance to reference (recovery; C). Note that the zero lines indicate that both levels have equal values. This means that hay meadows are closer to their reference than dry grasslands  (C)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_2_800dpi_16.5x5cm.tiff")
  )
```

@@

```{r fig.id = "figure-fcs", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The Favorable Conservation Status (FCS) of a dike experiment. The substrates had different sand admixtures and different seed mixture types were sown. The zero line indicate that target and non-target species are balanced. Positive values indicate that there are more target species.  Shown are the posterior distributions, their medians and 95 % credible intervals derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_4_fcs_epred_800dpi_16.5x8cm.tiff")
  )
```
